Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-07 paper/paper.qmd
%>% changed to |>

The remainder of this paper is structured as follows. [@sec-data] describes the data used in the analysis, including the sources and key variables and their visualizations.[@sec-model] outlines the models applied to estimate fire severity and identify key factors. [@sec-results] presents the results of the analysis. [@sec-discussion] provides an interpretation of the results in the context of current fire safety practices and suggests potential improvements. Finally, [@sec-appendixa] includes supplementary information and additional data details. [@sec-appendix-B] provided insight into Fire Incidence, about their sampling method and potential bias.

@@

This structured approach allows us to analyze and understand various aspects of fire incidents in Toronto. All variables in the dataset have the potential to provide valuable insights and perspectives for various analyses, contributing to fire prevention and safety improvements. However, for my analysis of fire incidents across Toronto, certain variables; such as latitude and longitude, incident numbers in the CAD system, Incident_Ward, and road intersections—are being excluded from the study. A more detailed explanation will be provided in the @sec-appendixa.

@@

**Understanding Variables of Interest**:\
To understand the distribution of various categorical and numerical variables in the fire incident dataset, I visualized each variable along with its distribution and reviewed its summary statistics. The goal is to gain insights into the key characteristics of the data, which will support the later stages of analysis and modeling. Can check @sec-appendixa for visualization/summary of the distribution of all variables. Below are a few important variables considered:

@@

2.  **Civilian Casualties**:\
    This is a numerical variable that provides insight into the severity of the incidents. Basic descriptive statistics (e.g., mean, median, and range) were calculated using the `summary()` function. Additionally, a scatter plot was created to explore the relationship between the number of civilian casualties and the incident ID (see @tbl-civilian-casualties-summary). We can see that most of the time, there are no casualties. This data set ranges across over 10 years (see @fig-fire-incidents-by-year in @sec-appendixa), looking at all levels of casualties, and taking its average over 6 years, we can see that this indicates the city has done a good job in protecting civilians from fire places.

@@

#heatmap
fire_data <- fire_data %>%
  mutate(
    loss_group = cut(
      estimated_dollar_loss,
      breaks = c(0, 1000, 10000, 50000, 500000, Inf),
      labels = c("<1k", "1k-10k", "10k-50k", "50k-500k", ">500k"),
      right = FALSE
    ),
    casualty_group = cut(
      total_casualties,
      breaks = c(0, 1, 2, 5, Inf),
      labels = c("0", "1", "2-5", ">5"),
      right = FALSE
    )
  )

severity_table <- fire_data %>%
  group_by(loss_group, casualty_group) %>%
  summarize(count = n())

@@

### Apparatus
fire_data_model <- fire_data_model %>%
  mutate(severity_bin = cut(number_of_responding_apparatus,
                            breaks = c(0, 10, 20, 30, 40, 50, Inf),
                            labels = c("[1,10]", "[11,20]", "[21,30]", "[31,40]", "[41,50]", ">50"),
                            right = TRUE))
severity_dist <- fire_data_model %>%
  group_by(severity_bin) %>%
  summarise(frequency = n()) %>%
  ungroup()
#severity_dist

@@

### Determining Severity
fire_data_model <- fire_data_model %>%
  mutate(
    # Define Casualty Severity
    casualty_severity = case_when(
      total_casualties == 0 ~ "Low", 
      total_casualties >= 4 ~ "High",
      TRUE ~ "Medium"
    ),

@@

      TRUE ~ "Low" 
    )
  )
severity_count_table <- fire_data_model %>%
  count(Severity)

@@

fire_data_model <- fire_data_model %>%
  mutate(Severity_high = as.factor(ifelse(Severity == "High", 1, 0)))

@@

fire_data_current <- fire_data_current %>%
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    fire_alarm_system_operation = factor(fire_alarm_system_operation) 
  )

top_fire_alarm_system_operation <- fire_data_current %>%
  group_by(Severity, fire_alarm_system_operation) %>%
  tally() %>%
  group_by(Severity) %>%
  top_n(5, n) %>%
  ungroup()

# Filter the dataset to include only the Operation of alarm for each severity
fire_operation_data_filtered <- fire_data_current %>%
  semi_join(top_fire_alarm_system_operation, by = c("Severity", "fire_alarm_system_operation"))

@@

# Convert Severity and sprinkler_system_data_filtered to factors
fire_data_current <- fire_data_current %>%
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    sprinkler_system_presence = factor(sprinkler_system_presence)  # Convert ignition_source_grouped to factor
  )

# Calculate the top 5 Status of Sprinkler System for each Severity group
top_sprinkler_system_presence <- fire_data_current %>%
  group_by(Severity, sprinkler_system_presence) %>%
  tally() %>%
  group_by(Severity) %>%
  top_n(5, n) %>%
  ungroup()

# Filter the dataset to include only the top Status of Sprinkler System for each severity
sprinkler_system_data_filtered <- fire_data_current %>%
  semi_join(top_sprinkler_system_presence, by = c("Severity", "sprinkler_system_presence"))

@@

#Convert Severity and area_of_origin_grouped to factors
fire_data_model <- fire_data_model %>%
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    area_of_origin_grouped = factor(area_of_origin_grouped)
  )

# Calculate the top 5 Area of Origins for each Severity group
top_area_of_origin <- fire_data_model %>%
  group_by(Severity, area_of_origin_grouped) %>%
  tally() %>%
  group_by(Severity) %>%
  top_n(5, n) %>%
  ungroup()

# Filter the dataset to include only the top Area of Origins for each severity
area_data_filtered <- fire_data_model %>%
  semi_join(top_area_of_origin, by = c("Severity", "area_of_origin_grouped"))

@@

## Factor of Ignition Source ##
# Convert Severity and ignition_source_grouped to factors
fire_data_model <- fire_data_model %>%
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    ignition_source_grouped = factor(ignition_source_grouped)  # Convert ignition_source_grouped to factor
  )

# Calculate the top 5 ignition sources for each Severity group
top_ignition_sources <- fire_data_model %>%
  group_by(Severity, ignition_source_grouped) %>%
  tally() %>%
  group_by(Severity) %>%
  top_n(5, n) %>%
  ungroup()

# Filter the dataset to include only the top ignition sources for each severity
fire_data_filtered <- fire_data_model %>%
  semi_join(top_ignition_sources, by = c("Severity", "ignition_source_grouped"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-possible-cause-full
#| tbl-cap: Comprehensive table showing all possible fire casues.
## 9. Possible Cause ##
#plot all that has frequency greater than 550
filtered_data <- fire_data %>%
  count(possible_cause) %>%
  filter(n > 550)
#Full Tabble
cause_table <- table(fire_data$possible_cause)
cause_table_df <- as.data.frame(cause_table) %>%
  arrange(desc(Freq))
kable(cause_table_df, col.names = c("Possible Cause", "Frequency"))
```

@@

# from year plot, we can see that 2022 and 2023 does not provide much information so we can droped them.
fire_data_filtered <- fire_data %>%
  filter(!is.na(year) & !is.na(month) & month >= 1 & month <= 12)
incident_count_by_year_month <- fire_data_filtered %>%
  group_by(year, month) %>%
  summarize(count = n(), .groups = 'drop')
incident_count_by_year_month_ex <- incident_count_by_year_month %>%
  filter(!(year %in% c(2022, 2023)))
incident_count_by_year_month_ex$month <- factor(incident_count_by_year_month_ex$month, levels = 1:12)
ggplot(incident_count_by_year_month_ex, aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "lightcoral", color = "black") +
  labs( x = "Month", y = "Incident Count") +
  scale_x_discrete(
    breaks = 1:12,  # Show all months (1 to 12)
    labels = month.name  # Use full month names
  ) +
  facet_wrap(~year, scales = "fixed") +  # Fix the y-axis scale for comparison
  theme_minimal() +
  theme(
    strip.text.x = element_text(size = 10), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.5, size = 8), 
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12), 
    strip.background = element_rect(fill = "lightblue", color = "black"), 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

>>>>

The remainder of this paper is structured as follows. [@sec-data] describes the data used in the analysis, including the sources and key variables and their visualizations.[@sec-model] outlines the models applied to estimate fire severity and identify key factors. [@sec-results] presents the results of the analysis. [@sec-discussion] provides an interpretation of the results in the context of current fire safety practices and suggests potential improvements. Finally, [@sec-appendix] includes supplementary information and additional data details.

@@

This structured approach allows us to analyze and understand various aspects of fire incidents in Toronto. All variables in the dataset have the potential to provide valuable insights and perspectives for various analyses, contributing to fire prevention and safety improvements. However, for my analysis of fire incidents across Toronto, certain variables; such as latitude and longitude, incident numbers in the CAD system, Incident_Ward, and road intersections—are being excluded from the study. A more detailed explanation will be provided in the @sec-appendix.

@@

**Understanding Variables of Interest**:\
To understand the distribution of various categorical and numerical variables in the fire incident dataset, I visualized each variable along with its distribution and reviewed its summary statistics. The goal is to gain insights into the key characteristics of the data, which will support the later stages of analysis and modeling. Can check @sec-appendix for visualization/summary of the distribution of all variables. Below are a few important variables considered:

@@

2.  **Civilian Casualties**:\
    This is a numerical variable that provides insight into the severity of the incidents. Basic descriptive statistics (e.g., mean, median, and range) were calculated using the `summary()` function. Additionally, a scatter plot was created to explore the relationship between the number of civilian casualties and the incident ID (see @tbl-civilian-casualties-summary). We can see that most of the time, there are no casualties. This data set ranges across over 10 years (see @fig-fire-incidents-by-year in @sec-appendix), looking at all levels of casualties, and taking its average over 6 years, we can see that this indicates the city has done a good job in protecting civilians from fire places.

@@

#heatmap
fire_data <- fire_data |>
  mutate(
    loss_group = cut(
      estimated_dollar_loss,
      breaks = c(0, 1000, 10000, 50000, 500000, Inf),
      labels = c("<1k", "1k-10k", "10k-50k", "50k-500k", ">500k"),
      right = FALSE
    ),
    casualty_group = cut(
      total_casualties,
      breaks = c(0, 1, 2, 5, Inf),
      labels = c("0", "1", "2-5", ">5"),
      right = FALSE
    )
  )

severity_table <- fire_data |>
  group_by(loss_group, casualty_group) |>
  summarize(count = n())

@@

### Apparatus
fire_data_model <- fire_data_model |>
  mutate(severity_bin = cut(number_of_responding_apparatus,
                            breaks = c(0, 10, 20, 30, 40, 50, Inf),
                            labels = c("[1,10]", "[11,20]", "[21,30]", "[31,40]", "[41,50]", ">50"),
                            right = TRUE))
severity_dist <- fire_data_model |>
  group_by(severity_bin) |>
  summarise(frequency = n()) |>
  ungroup()
#severity_dist

@@

### Determining Severity
fire_data_model <- fire_data_model |>
  mutate(
    # Define Casualty Severity
    casualty_severity = case_when(
      total_casualties == 0 ~ "Low", 
      total_casualties >= 4 ~ "High",
      TRUE ~ "Medium"
    ),

@@

      TRUE ~ "Low" 
    )
  )
severity_count_table <- fire_data_model |>
  count(Severity)

@@

fire_data_model <- fire_data_model |>
  mutate(Severity_high = as.factor(ifelse(Severity == "High", 1, 0)))

@@

fire_data_current <- fire_data_current |>
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    fire_alarm_system_operation = factor(fire_alarm_system_operation) 
  )

top_fire_alarm_system_operation <- fire_data_current |>
  group_by(Severity, fire_alarm_system_operation) |>
  tally() |>
  group_by(Severity) |>
  top_n(5, n) |>
  ungroup()

# Filter the dataset to include only the Operation of alarm for each severity
fire_operation_data_filtered <- fire_data_current |>
  semi_join(top_fire_alarm_system_operation, by = c("Severity", "fire_alarm_system_operation"))

@@

# Convert Severity and sprinkler_system_data_filtered to factors
fire_data_current <- fire_data_current |>
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    sprinkler_system_presence = factor(sprinkler_system_presence)  # Convert ignition_source_grouped to factor
  )

# Calculate the top 5 Status of Sprinkler System for each Severity group
top_sprinkler_system_presence <- fire_data_current |>
  group_by(Severity, sprinkler_system_presence) |>
  tally() |>
  group_by(Severity) |>
  top_n(5, n) |>
  ungroup()

# Filter the dataset to include only the top Status of Sprinkler System for each severity
sprinkler_system_data_filtered <- fire_data_current |>
  semi_join(top_sprinkler_system_presence, by = c("Severity", "sprinkler_system_presence"))

@@

#Convert Severity and area_of_origin_grouped to factors
fire_data_model <- fire_data_model |>
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    area_of_origin_grouped = factor(area_of_origin_grouped)
  )

# Calculate the top 5 Area of Origins for each Severity group
top_area_of_origin <- fire_data_model |>
  group_by(Severity, area_of_origin_grouped) |>
  tally() |>
  group_by(Severity) |>
  top_n(5, n) |>
  ungroup()

# Filter the dataset to include only the top Area of Origins for each severity
area_data_filtered <- fire_data_model |>
  semi_join(top_area_of_origin, by = c("Severity", "area_of_origin_grouped"))

@@

## Factor of Ignition Source ##
# Convert Severity and ignition_source_grouped to factors
fire_data_model <- fire_data_model |>
  mutate(
    Severity = factor(Severity, levels = c("Low", "Medium", "High")),
    ignition_source_grouped = factor(ignition_source_grouped)  # Convert ignition_source_grouped to factor
  )

# Calculate the top 5 ignition sources for each Severity group
top_ignition_sources <- fire_data_model |>
  group_by(Severity, ignition_source_grouped) |>
  tally() |>
  group_by(Severity) |>
  top_n(5, n) |>
  ungroup()

# Filter the dataset to include only the top ignition sources for each severity
fire_data_filtered <- fire_data_model |>
  semi_join(top_ignition_sources, by = c("Severity", "ignition_source_grouped"))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-possible-cause-full
#| tbl-cap: Comprehensive table showing all possible fire casues.
## 9. Possible Cause ##
#plot all that has frequency greater than 550
filtered_data <- fire_data |>
  count(possible_cause) |>
  filter(n > 550)
#Full Tabble
cause_table <- table(fire_data$possible_cause)
cause_table_df <- as.data.frame(cause_table) |>
  arrange(desc(Freq))
kable(cause_table_df, col.names = c("Possible Cause", "Frequency"))
```

@@

# from year plot, we can see that 2022 and 2023 does not provide much information so we can droped them.
fire_data_filtered <- fire_data |>
  filter(!is.na(year) & !is.na(month) & month >= 1 & month <= 12)
incident_count_by_year_month <- fire_data_filtered |>
  group_by(year, month) |>
  summarize(count = n(), .groups = 'drop')
incident_count_by_year_month_ex <- incident_count_by_year_month |>
  filter(!(year %in% c(2022, 2023)))
incident_count_by_year_month_ex$month <- factor(incident_count_by_year_month_ex$month, levels = 1:12)
ggplot(incident_count_by_year_month_ex, aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "lightcoral", color = "black") +
  labs( x = "Month", y = "Incident Count") +
  scale_x_discrete(
    breaks = 1:12,  # Show all months (1 to 12)
    labels = month.name  # Use full month names
  ) +
  facet_wrap(~year, scales = "fixed") +  # Fix the y-axis scale for comparison
  theme_minimal() +
  theme(
    strip.text.x = element_text(size = 10), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.5, size = 8), 
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12), 
    strip.background = element_rect(fill = "lightblue", color = "black"), 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )
